---
title: "PregnancyIdentifier pipeline overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PregnancyIdentifier pipeline overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

This vignette describes the **PregnancyIdentifier** pipeline, which implements the HIPPS approach to identify pregnancy episodes in OMOP CDM data. The pipeline has four main algorithm components, run in sequence:

1.  **HIP** — Outcome-anchored episode builder: identifies pregnancy episodes from high-specificity pregnancy outcomes and gestational-age evidence. See the [HIP vignette](hip.html).
2.  **PPS** — Gestational-timing concept episode builder: constructs episodes from pregnancy-related concepts and expected timing windows. See the [PPS vignette](pps.html).
3.  **Merge** — Combines HIP and PPS episodes into a unified per-person timeline (HIPPS) by temporal overlap and best-match deduplication. See the [Merge vignette](merge.html).
4.  **ESD** — Episode start date refinement: uses gestational timing evidence (GW and GR3m) to infer pregnancy start dates and precision. See the [ESD vignette](esd.html).

## Running the full pipeline

The primary entry point is `runPregnancyIdentifier()`, which runs all steps end-to-end and writes intermediate artifacts and final outputs to `outputDir`.

```{r, eval=FALSE}
library(PregnancyIdentifier)
library(CDMConnector)

# A cdm_reference from CDMConnector (e.g. mock or real CDM)
cdm <- mockPregnancyCdm()

runPregnancyIdentifier(
  cdm                   = cdm,
  outputDir             = "pregnancy_identifier_output",
  startDate             = as.Date("2000-01-01"),
  endDate               = Sys.Date(),
  justGestation         = TRUE,
  minCellCount          = 5L,
  runExport             = FALSE,
  conformToValidation   = FALSE
)
```

### Outputs written to disk

The pipeline writes intermediate RDS artifacts as it proceeds and finishes with a patient-level episode table plus optional shareable exports:

| File | Description |
|----------------------|--------------------------------------------------|
| `runStart.csv` | Run timestamp used in exported csv files |
| `log.txt` | Run log |
| `hip_episodes.rds` | HIP episodes (outcome and/or gestation-derived) |
| `pps_episodes.rds` | PPS episodes with inferred outcomes (input to merge) |
| `pps_concept_counts.csv` | PPS concept record counts per concept (used by export when `runExport = TRUE`). |
| `pps_gest_timing_episodes.rds` | (Only when `debugMode = TRUE`) Record-level PPS concept rows with episode assignment. |
| `pps_min_max_episodes.rds` | (Only when `debugMode = TRUE`) Episode-level PPS min/max date summaries. |
| `hipps_episodes.rds` | Merged HIP + PPS episode table (standardized columns) |
| `final_pregnancy_episodes.rds` | **Final** episode table: one row per episode, with inferred start/end and outcomes |
| `esd.rds` | (Only when `debugMode = TRUE`) ESD episode-level start inference before merging back to HIPPS. |
| `export/` | De-identified summary CSVs and ZIP archive when `runExport = TRUE`. See the [Export vignette](export.html) for file names, columns, and analysis use. |

## Pipeline steps in order

1.  **initPregnancies** — Loads pregnancy concept sets into the CDM, builds `preg_hip_records` and `preg_pps_records` from condition, procedure, observation, and measurement tables within the study window.
2.  **runHip** — Builds outcome episodes, then gestation episodes, merges and cleans them, and writes `hip_episodes.rds`.
3.  **runPps** — Builds PPS episodes from pregnancy-related concepts, assigns outcomes, and writes `pps_episodes.rds`.
4.  **mergeHipps** — Reads `hip_episodes.rds` and `pps_episodes.rds`, merges by temporal overlap, deduplicates many-to-many matches, and writes `hipps_episodes.rds`.
5.  **runEsd** — Reads `hipps_episodes.rds`, pulls gestational timing concepts (GW/GR3m), infers start dates and precision, harmonizes end and outcome from HIP vs PPS, and writes `final_pregnancy_episodes.rds`. When `conformToValidation = TRUE`, episode output is modified to remove overlapping episodes and episodes longer than 308 days.
6.  **exportPregnancies** — (Optional, when `runExport = TRUE`) Reads `final_pregnancy_episodes.rds` and writes shareable CSVs and a ZIP to `file.path(outputDir, "export")`. See the [Export vignette](export.html).

## What each step contributes to final start, end, and outcome

The pipeline’s most important analysis variables are **final_episode_start_date**, **final_episode_end_date**, and **final_outcome_category**. Here is what each step contributes to them.

| Variable | Step | Contribution |
|----------|------|--------------|
| **Final start** | **initPregnancies** | Supplies raw pregnancy-related concepts used later by HIP, PPS, and ESD. |
| | **HIP** | Produces estimated start (e.g. from outcome + term range); used in merge interval and as context for ESD. |
| | **PPS** | Produces episode bounds (min/max dates) from gestational timing; used in merge interval. |
| | **Merge** | Produces **merge_episode_start** / **merge_episode_end**, which define the **evidence window** ESD uses to pull GW/GR3m concepts. Does *not* set the final start. |
| | **ESD** | **Sets final start.** Infers **final_episode_start_date** from GW/GR3m concepts within the merged window; if no start is inferred, uses final end − max_term (Matcho term for the chosen outcome). See [ESD vignette](esd.html). |
| **Final end** | **HIP** | Produces **hip_end_date** (outcome or gestation end). |
| | **PPS** | Produces **pps_end_date** (outcome or episode max date). |
| | **Merge** | Puts HIP and PPS on one row per episode; produces **merge_episode_end** (used only as evidence window by ESD). Does *not* set the final end. |
| | **ESD** | **Sets final end.** Chooses **final_episode_end_date** by picking between **hip_end_date** and **pps_end_date** using harmonization rules (e.g. outcome match, or “take the one that occurs second”). Does *not* use **merge_episode_end** for the reported end. See [ESD vignette](esd.html). |
| **Final outcome** | **HIP** | Produces **hip_outcome_category**. |
| | **PPS** | Produces **pps_outcome_category**. |
| | **Merge** | Keeps both **hip_outcome_category** and **pps_outcome_category** on each row. Does *not* choose a single outcome. |
| | **ESD** | **Sets final outcome.** Chooses **final_outcome_category** by picking between **hip_outcome_category** and **pps_outcome_category** using the same rules as for the end date. See [ESD vignette](esd.html). |

So: **final start** comes from ESD (GW/GR3m inference or end − max_term). **Final end** and **final outcome** come from ESD’s choice between HIP and PPS; the merge only puts both values on the same row so ESD can pick one.

## Interpreting the final episode table

The primary analysis dataset is `final_pregnancy_episodes.rds`, with one row per pregnancy episode. Key fields include:

- **final_episode_start_date** — Best estimate of pregnancy start for this episode. Set by ESD: when ESD has gestational timing (GW and/or GR3m) within the merged evidence window, the start is inferred from that (e.g. concept_date − 7×weeks for GW, or GR3m midpoint). When ESD has no inferred start, it is set to **final_episode_end_date − max_term** (Matcho term for the chosen outcome, e.g. 301 days for live birth).
- **final_episode_end_date** — Best estimate of pregnancy end for this episode. **Not** taken from the merged interval (**merge_episode_end**). Set by ESD by **picking one of** **hip_end_date** or **pps_end_date**: when HIP and PPS agree (same outcome and end within 14 days), ESD uses hip_end_date; when one side is missing, ESD uses the side present; when they disagree, ESD uses rules (e.g. take the one that occurs second, or HIP when timing is similar). See the [ESD vignette](esd.html) for the full logic.
- **final_outcome_category** — Outcome category for this episode (e.g. LB, SB, PREG). Set by ESD by **picking one of** **hip_outcome_category** or **pps_outcome_category** using the same priority rules as for the end date.

Together, **final_episode_start_date**, **final_episode_end_date**, and **final_outcome_category** are the primary analysis variables: the pipeline’s final inferred pregnancy interval and outcome.

- **merge_episode_start** — Start of the evidence window for this episode from the merge step. It is the minimum of: HIP’s first gestation date in the episode, PPS episode min date, and HIP estimated start. So it is the earliest date implied by either algorithm for this merged episode.
- **merge_episode_end** — End of the evidence window for this episode from the merge step. It is the maximum of: PPS episode max date and HIP pregnancy end (visit/outcome date). So it is the latest date implied by either algorithm for this merged episode.

**merge_episode_start** and **merge_episode_end** describe the span of raw evidence (HIP + PPS) that was merged for this episode, before ESD refines start/end into the inferred interval.

- **hip_end_date** — Episode end date from the HIP algorithm: either the outcome date (e.g. delivery) or the latest gestation date when there is no outcome. **NA** when the episode was found only by PPS (no overlapping HIP episode).
- **pps_end_date** — Episode end from the PPS algorithm: the inferred outcome date when PPS assigned an outcome (LB, SB, etc.), otherwise the PPS episode max date. **NA** when the episode was found only by HIP (no overlapping PPS episode).

**hip_end_date** and **pps_end_date** are the end dates produced by each algorithm; at least one is non-NA when both algorithms contributed to the episode, and one is NA for one-sided (HIP-only or PPS-only) episodes.

## An example case

```{r}
library(PregnancyIdentifier)
library(CDMConnector)
library(dplyr, warn.conflicts = FALSE)

cdm <- mockPregnancyCdm()

cdm <- cdmSubset(cdm, personId = 24)

cdm %>% 
  cdmFlatten() %>% 
  select(-"type_concept_id", -"domain") %>% 
  arrange(start_date)

```

```{r, cache=TRUE}
outputDir <- file.path(tempdir(), "pregnancy_output")

invisible(capture.output(
  runPregnancyIdentifier(cdm, outputDir = outputDir)
))

list.files(outputDir)
readRDS(file.path(outputDir, "final_pregnancy_episodes.rds"))
```


