---
title: "Export: shareable summary CSVs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Export: shareable summary CSVs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
# Ensure tempdir() exists (avoids build failures when TMPDIR is set but missing)
if (!dir.exists(tempdir())) dir.create(tempdir(), recursive = TRUE, showWarnings = FALSE)
```

# Overview

The pipeline writes **person-level and episode-level data** (e.g. `final_pregnancy_episodes.rds`) to `outputFolder`. The **export** step turns those results and related artifacts into **shareable aggregated CSV files** for QA, cross-site comparison, and use as input to the Shiny app. Export runs automatically as part of `runPregnancyIdentifier()` and writes to `exportFolder` (default `file.path(outputFolder, "export")`). Small cell counts can be suppressed using `minCellCount`. Alternatively, run `exportPregnancies()` yourself to write to a chosen directory.

This vignette lists every exported CSV: file name, columns, meaning of each column, and how each file is used in analysis.

# How to run export

Export runs automatically in `runPregnancyIdentifier()`. Shareable CSVs go to `exportFolder`; default is `file.path(outputFolder, "export")`. Override with the `exportFolder` argument:

```{r, eval=FALSE}
library(PregnancyIdentifier)
cdm <- mockPregnancyCdm()
# Export runs by default; CSVs go to "out/export":
runPregnancyIdentifier(cdm, outputFolder = "out")
# Or write export CSVs to a custom folder:
runPregnancyIdentifier(cdm, outputFolder = "out", exportFolder = "path/to/my_export")
```

To run export alone (e.g. after a previous pipeline run), use `exportPregnancies()` with the same `outputFolder` and your chosen `exportFolder`. The same CSVs are produced; `exportFolder` in `runPregnancyIdentifier()` corresponds to `exportFolder` in `exportPregnancies()`.

---

# Exported files: metadata and copied artifacts

## cdm_source.csv

**Source:** `CDMConnector::snapshot(cdm)` — one row of CDM metadata.

**Columns:** Depend on the CDMConnector version; typically include identifiers and timestamps for the CDM (e.g. `cdm_name`, `snapshot_date`, and other snapshot fields).

**Purpose:** Documents which CDM and snapshot the export came from. Used when comparing or aggregating exports across sites or runs so you can track provenance and CDM version.

---

## pps_concept_counts.csv

**Source:** Copied from `outputFolder` (written by `runPps()`).

**Content:** Counts of pregnancy-related concepts used by PPS per concept (`pps_concept_id`, `pps_concept_name`, `record_count`, `person_count`).

**Purpose:** Describes the PPS concept usage in this run. Used to compare concept coverage across sites and to check that expected pregnancy concepts are present.

---

## hip_concept_counts.csv

**Source:** Copied from `outputFolder` when present (written by the pipeline when HIP concept counts are produced).

**Content:** Counts of HIP pregnancy-related concepts per concept.

**Purpose:** Describes HIP concept usage in this run; used for concept coverage and cross-site comparison.

---

## esd_concept_counts.csv

**Source:** Copied from `outputFolder` when present (written by the pipeline when ESD concept counts are produced).

**Content:** Counts of ESD timing concepts used per concept.

**Purpose:** Describes ESD concept usage; used for QA and cross-site comparison.

---

## attrition.csv

**Source:** Copied from `outputFolder` when present (written by pipeline steps such as HIP and merge).

**Content:** Step-by-step record and person counts (prior/post, dropped) through the pipeline.

**Purpose:** Audit trail of attrition at each step; used for debugging and cohort flow.

---

## log.txt

**Source:** Copied from `outputFolder` (pipeline log from `makeLogger()`).

**Content:** Plain-text log of pipeline steps (init, HIP, PPS, merge, ESD).

**Purpose:** Audit trail for the run. Used for debugging and to verify which steps ran and when.

---

# Exported files: derived from final_pregnancy_episodes.rds

All of the following are computed from `final_pregnancy_episodes.rds` and share common metadata columns when present: **cdm_name**, **date_run** (pipeline run timestamp), **date_export** (export/snapshot date), **pkg_version** (PregnancyIdentifier version). Counts in the range `(0, minCellCount)` may be replaced with `NA` for disclosure control.

---

## age_summary.csv

**Columns:** Summary of **age_pregnancy_start** (age in years at pregnancy start): **final_outcome_category**, **n**, **min**, **Q25**, **median**, **Q75**, **max**, **mean**, plus metadata.

**Purpose:** Describes the distribution of maternal age at pregnancy start. Used for cohort description, feasibility checks, and comparing age distributions across sites or time windows.

---

## age_summary_groups.csv

**Columns:** **colName**, **age_pregnancy_start** (value or group: integer year or `"<12"`, `">55"`), **n**, **total**, **pct**, plus metadata. Counts may be suppressed by `minCellCount`.

**Purpose:** Counts and percentages of pregnancies by age (by year and by boundary groups &lt;12 and &gt;55). Used for age-stratified summaries and checking extreme-age pregnancy counts.

---

## age_summary_first_pregnancy.csv

**Columns:** **final_outcome_category**, **colName**, **min**, **Q25**, **median**, **Q75**, **max**, **mean**, **sd**, plus metadata. One row per outcome group: **overall** (any outcome), each individual outcome (LB, SB, AB, SA, ECT, DELIV, PREG), and **LB or PREG** (combined).

**Purpose:** Age at first pregnancy start, stratified by outcome; used for cohort description and feasibility.

---

## age_summary_first_pregnancy_end.csv

**Columns:** **final_outcome_category**, **colName**, **min**, **Q25**, **median**, **Q75**, **max**, **mean**, **sd**, plus metadata. Same structure as `age_summary_first_pregnancy.csv` but summarises **age_pregnancy_end** (age at pregnancy end date) instead of age at start.

**Purpose:** Age at first pregnancy end, stratified by outcome; used for cohort description and feasibility.

---

## precision_days.csv

**Columns:** **esd_precision_days** (x-axis values), **density** (kernel density estimate), plus **cdm_name**, **date_run**, **date_export**, **pkg_version**.

**Purpose:** Distribution of ESD start-date precision (in days). Used to see how precise inferred start dates are (e.g. mostly week-level vs month-level) and to compare precision across sites.

---

## precision_days_denominators.csv

**Columns:** **total_episodes**, **episodes_with_precision_days**, **pct_with_precision_days**, **episodes_with_gw_timing**, **pct_with_gw_timing**, plus metadata.

**Purpose:** Denominators for interpreting precision outputs (how many episodes had ESD precision and GW timing evidence).

---

## episode_frequency.csv

**Columns:** **total_episodes**, **total_individuals** (distinct persons with at least one episode), plus metadata. **total_individuals** may be suppressed if &lt; minCellCount.

**Purpose:** High-level counts: how many pregnancy episodes and how many people. Used as the main denominator for rates and for site-level summaries.

---

## pregnancy_frequency.csv

**Columns:** **freq** (number of pregnancies per person: 1, 2, 3, …), **number_individuals** (count of persons with that frequency), plus metadata. **number_individuals** may be suppressed.

**Purpose:** Distribution of pregnancy count per person (parity-like). Used to describe repeat pregnancies and to check for implausible multiplicity.

---

## episode_frequency_summary.csv

**Columns:** Summary of **freq** (episodes per person): **colName**, **min**, **Q25**, **median**, **Q75**, **max**, **mean**, **sd**, plus metadata.

**Purpose:** Numeric summary of how many episodes per person (min/median/max, etc.). Used for cohort description and cross-site comparison of pregnancy frequency.

---

## gestational_age_days_summary.csv

**Columns:** Summary of **esd_gestational_age_days_calculated** (gestational length in days): **colName**, **min**, **Q25**, **median**, **Q75**, **max**, **mean**, **sd**, plus metadata.

**Purpose:** Distribution of gestational duration. Used to describe pregnancy length, flag implausible durations, and compare sites.

---

## gestational_age_days_counts.csv

**Columns:** **less_1day** (count of episodes with gestational age &lt; 1 day), **over_308days** (count &gt; 308 days), plus metadata.

**Purpose:** Counts of out-of-range gestational ages. Used for data quality and to quantify implausible or missing duration.

---

## gestational_weeks.csv

**Columns:** **final_outcome_category**, **gestational_weeks** (integer: floor of gestational age in days / 7), **n** (number of episodes), **pct** (percentage), plus metadata.

**Purpose:** Distribution of gestational age by week. Used for gestational-age histograms, preterm/term summaries, and cross-site comparison.

---

## gestational_age_days_per_category_summary.csv

**Columns:** One row per **final_outcome_category** (e.g. LB, SB, PREG). For each: summary of **esd_gestational_age_days_calculated** — **colName**, **min**, **Q25**, **median**, **Q75**, **max**, **mean**, **sd**, plus **final_outcome_category** and metadata.

**Purpose:** Gestational duration by outcome type. Used to check that outcome-specific durations (e.g. live birth vs miscarriage) are plausible and to compare across sites.

---

## yearly_trend.csv

**Columns:** **column** (name of the date field), **year** (integer), **count** (number of episodes with that date in that year), plus metadata.

**Purpose:** Episode counts by year for each date type (e.g. final_episode_start_date, final_episode_end_date). Used for temporal trends and study window checks.

---

## yearly_trend_missing.csv

**Columns:** Same as yearly_trend but rows where **year** is NA (missing date).

**Purpose:** Count of episodes with missing dates by column. Used for completeness and missing-data reporting.

---

## monthly_trends.csv

**Columns:** **column** (date field), **month** (month name), **count**, plus metadata.

**Purpose:** Episode counts by month (seasonality). Used for seasonal patterns and data completeness by month.

---

## monthly_trend_missing.csv

**Columns:** Same as monthly_trends but rows where **month** is NA.

**Purpose:** Count of episodes with missing date by column, for completeness.

---

## observation_period_range.csv

**Columns:** **min_obs** (minimum observation period start year), **max_obs** (maximum observation period end year), plus metadata. Derived from CDM `observation_period`.

**Purpose:** Range of observation periods in the CDM. Used to interpret temporal coverage and to compare study windows across sites.

---

## pregnancy_overlap_counts.csv

**Columns:** **colName** (value `"overlap"` for all rows), **overlap** (FALSE or TRUE per row), **n** (count of records in that category), **total** (total episode records, same for all rows), **pct** (percentage of records in that category), plus **cdm_name**, **date_run**, **date_export**, **pkg_version**. One row per overlap category: FALSE (episode does not overlap another—no temporal overlap with previous episode, or no previous episode for that person, e.g. first episode), TRUE (episode overlaps another: episode start ≤ previous episode end within person).

**Purpose:** Summary counts of overlapping inferred pregnancy intervals. Used for data quality (overlaps may indicate algorithm or data issues).

---

## missing_dates.csv

**Columns:** Number and percentage of missing start and end dates for HIP episodes, PPS episodes, and ESD (final) episodes: **hip_start_n**, **hip_start_pct**, **hip_end_n**, **hip_end_pct**, **pps_start_n**, **pps_start_pct**, **pps_end_n**, **pps_end_pct**, **esd_start_n**, **esd_start_pct**, **esd_end_n**, **esd_end_pct**, plus **cdm_name**, **date_run**, **date_export**, **pkg_version**. HIP uses merge_pregnancy_start (start) and hip_end_date (end); PPS uses pps_episode_min_date (start) and pps_episode_max_date (end); ESD uses final_episode_start_date and final_episode_end_date.

**Purpose:** Counts and percentages of missing start/end dates by episode type. Used to assess completeness of key dates and to compare across sites.

---

## swapped_dates.csv

**Columns:** **source** (hip, pps, or esd), **n_swapped** (count where start &gt; end), **total** (records with both dates for that source), **pct** (100 × n_swapped / total), plus **cdm_name**, **date_run**, **date_export**, **pkg_version**. One row per source: HIP (merge_pregnancy_start &gt; hip_end_date), PPS (merge_pregnancy_start &gt; pps_end_date), ESD (final_episode_start_date &gt; final_episode_end_date).

**Purpose:** Count of episodes with reversed (swapped) start/end by source. Used for data quality and algorithm validation.

---

## outcome_categories_count.csv

**Columns:** **outcome_category** (e.g. LB, SB, PREG), **algorithm** (hip, pps, hipps), **n**, **pct** (within algorithm), plus metadata.

**Purpose:** Outcome distribution by algorithm (HIP, PPS, final harmonized). Used to compare algorithm agreement and to describe outcome mix across sites.

---

## delivery_mode_summary.csv

**Columns:** **final_outcome_category**, delivery-mode counts (e.g. cesarean, vaginal), **n**, plus metadata.

**Purpose:** Delivery mode (cesarean vs vaginal) by outcome category for live-birth analyses.

---

## concept_check.csv

**Columns:** Concept-timing check results (concept coverage and timing vs episodes), plus metadata.

**Purpose:** QA of concept usage and timing relative to pregnancy episodes.

---

## quality_check_cleanup.csv

**Columns:** Quality metrics for cleanup/validation (e.g. overlap counts, episode length flags), plus metadata.

**Purpose:** QA of pipeline cleanup and validation steps.

---

## attrition_if_cleanup.csv

**Columns:** Counts of episodes and people flagged by each cleanup rule and retained after cleanup checks, plus metadata.

**Purpose:** Documents cleanup quality checks and potential attrition under cleanup rules.

---

# ZIP archive

`exportPregnancies()` writes CSVs only. To create a ZIP archive, run `zipExportFolder(exportFolder)` after export (and after PET comparison outputs if you want one combined archive).

# Summary table

| File | Main content | Main use in analysis |
|------|--------------|----------------------|
| cdm_source.csv | CDM/snapshot metadata | Provenance, site/CDM identification |
| pps_concept_counts.csv | PPS concept counts | Concept coverage, PPS input description |
| hip_concept_counts.csv | HIP concept counts | HIP concept coverage (when present) |
| esd_concept_counts.csv | ESD concept counts | ESD concept coverage (when present) |
| attrition.csv | Pipeline step attrition | Audit, cohort flow (when present) |
| log.txt | Pipeline log | Audit, debugging |
| age_summary.csv | Age-at-start distribution (summary stats) | Cohort description, feasibility |
| age_summary_first_pregnancy.csv | Age at first pregnancy start by outcome | First-pregnancy age description |
| age_summary_first_pregnancy_end.csv | Age at first pregnancy end by outcome | First-pregnancy end age description |
| age_summary_groups.csv | Age counts (by year and &lt;12, &gt;55) | Age stratification, boundary checks |
| precision_days.csv | Density of ESD precision (days) | Start-date precision, site comparison |
| precision_days_denominators.csv | Precision/GW timing denominators | Context for precision completeness |
| episode_frequency.csv | Total episodes, total individuals | Denominators, site summaries |
| pregnancy_frequency.csv | Episodes per person (1, 2, 3, …) | Parity-like distribution |
| episode_frequency_summary.csv | Summary of episodes per person | Cohort description |
| gestational_age_days_summary.csv | Summary of gestational length (days) | Duration distribution |
| gestational_age_days_counts.csv | Counts &lt;1 day, &gt;308 days | Out-of-range duration QA |
| gestational_weeks.csv | Counts by gestational week | GA distribution, preterm/term |
| gestational_age_days_per_category_summary.csv | Gestational length by outcome | Outcome-specific duration |
| yearly_trend.csv | Episode count by year and date column | Temporal trends |
| yearly_trend_missing.csv | Missing-date count by column/year | Completeness |
| monthly_trends.csv | Episode count by month | Seasonality, completeness |
| monthly_trend_missing.csv | Missing-date count by column/month | Completeness |
| observation_period_range.csv | Min/max observation period years | Study window, coverage |
| pregnancy_overlap_counts.csv | Overlap summary (n_overlap_true/false, etc.) | Overlap QA |
| missing_dates.csv | N and % missing start/end (HIP, PPS, ESD) | Completeness by episode type |
| swapped_dates.csv | Count of start &gt; end (HIP/PPS/ESD) | Date logic QA |
| outcome_categories_count.csv | Outcome counts by algorithm | Outcome mix, algorithm agreement |
| delivery_mode_summary.csv | Delivery mode by outcome | Cesarean/vaginal by outcome |
| concept_check.csv | Concept timing vs episodes | Concept QA |
| quality_check_cleanup.csv | Cleanup/validation QA | Cleanup QA |
| attrition_if_cleanup.csv | Cleanup-rule flagged counts | Cleanup QA and potential attrition |
